---
title: "NEWS score calculation"
output: rmarkdown::html_vignette
bibliography: bibliography.bib
link-citations: true
reference-section-title: References
csl: ama-brackets.csl
vignette: >
  %\VignetteIndexEntry{NEWS score calculation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, eval = FALSE, include = FALSE}
library(acutelines.datatoolbox)
```

**DISCLAMER:** This function calculates the original National Early Warning Score (NEWS) [@News2012], not NEWS-2 [@News2_2017]. The scoring rules differ slightly from NEWS-2, specifically in oxygen saturation thresholds and mental status classification.

# Usage
The NEWS can be calculated using the `news_total()` function. This function requires vital parameters as input, which are then scored according to the NEWS criteria.
By default `return_df` is `FALSE` and can be left out, if set to `TRUE` the function will return all the inidivual elements of the NEWS score instead of only the totals.
Assuming your vital parameters are stored in dataframe `df` with columns `hr`, `rf`, `temp`, `sbp`, `avpu`, `gcs`, `o2supp`, and `spo2`, you can calculate the NEWS score as follows:

``` R
result <- news_total(df$hr, df$rf, df$temp, df$sbp, df$avpu, df$gcs, df$o2supp, df$spo2)
```

By default, the function only returns the total NEWS score, but you can also return the individual components of the NEWS score by setting `return_df = TRUE`. This will return a dataframe with all the individual components of the NEWS score.

``` R
result <- news_total(df$hr, df$rf, df$temp, df$sbp, df$avpu, df$gcs, df$o2supp, df$spo2, return_df = TRUE)
```

If all eight components scores are missing, the returned total NEWS score will be `NA`.

# Calculations

## Heart rate (HR)

|     Points    |     Criteria (bpm)     |
|---------------|------------------------|
|     0         |     51–90              |
|     1         |     41–50 OR 91–110    |
|     2         |     111–130            |
|     3         |     ≤ 40 OR ≥ 131      |

If HR is missing, `NA` is returned.

## Respiratory rate (RR)

|     Points    |     Criteria (breaths/min)    |
|---------------|-------------------------------|
|     0         |     12–20                     |
|     1         |     9–11                      |
|     2         |     21–24                     |
|     3         |     ≤ 8 OR ≥ 25               |

If RR is missing, `NA` is returned.

## Temperature (Temp)

|     Points    |     Criteria (°C)             |
|---------------|-------------------------------|
|     0         |     36.1–38.0                 |
|     1         |     35.1–36.0 OR 38.1–39.0    |
|     2         |     ≥ 39.1                    |
|     3         |     ≤ 35.0                    |

If Temp is missing, `NA` is returned.

## Systolic blood pressure (SBP)

|     Points    |     Criteria (mmHg)    |
|---------------|------------------------|
|     0         |     111–219            |
|     1         |     101–110            |
|     2         |     91–100             |
|     3         |     ≤ 90 OR ≥ 220      |

If SBP is missing, `NA` is returned.

## Level of consciousness (AVPU/GCS)

Variables used:

* AVPU (Alert, Voice, Pain, Unresponsive, or “D” for “Drowsy”).
* GCS (Glasgow Coma Scale).

NOTE: AVPU is assessed first. Only if AVPU is missing is GCS used as a substitute.

|     Points    |     Criteria                                            |
|---------------|---------------------------------------------------------|
|     0         |     AVPU = Alert (A) OR GCS = 15                        |
|     3         |     AVPU = V, P, U, D OR (AVPU missing AND GCS ≤ 14)    |

If both AVPU and GCS are missing, `NA` is returned.

## Oxygen supplementation (O2supp)

|     Points    |     Criteria (binary)                             |
|---------------|---------------------------------------------------|
|     0         |     No O~2~ supplementation   (value = FALSE)     |
|     2         |     O~2~ supplementation present (value = TRUE)   |

If O2supp is missing, `NA` is returned.

## Oxygen saturation (SpO~2~)

|     Points    |     Criteria    |
|---------------|-----------------|
|     0         |     ≥ 96%       |
|     1         |     94–95%      |
|     2         |     92–93%      |
|     3         |     ≤ 91%       |

If SpO~2~ is missing, `NA` is returned.

