---
title: "SOFA score calculation"
output: rmarkdown::html_vignette
bibliography: bibliography.bib
link-citations: true
reference-section-title: References
csl: ama-brackets.csl
vignette: >
  %\VignetteIndexEntry{SOFA score calculation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, eval = FALSE, include = FALSE}
library(acutelines.datatoolbox)
```

# Usage
Asuming the data is loaded into a dataframe `df`, the SOFA score can be calculated in two different ways.

## Automatic calculation
Use the `sofa_magic_wrapper()` function when you use a raw data file from the Acutelines database. SOFA related variables are coded as `SOFA_[timespan]h_[variable]`. Make sure all variables are present.
`interval` can be used to set the time stap interval in correspondance to your dataset, while `timespan` can be use to set the maximum timespan. By default `return_df` is `FALSE` and can be left out, if
set to `TRUE` the function will return all the inidivual elements of the SOFA scora instead of only the totals.

``` R
result <- sofa_magic_wrapper(df, interval=24, timespan=72)
```

<details>
<summary>Manual calculation</summary>
When your dataset has a different naming schema (not having the `SOFA_[timespan]h_[variable]` format), you can still calculate the SOFA score with this package, however, you have to
provide the package with a mapping of each column in your dataset corresponding to the acutal element of the SOFA score.

First define the mapping:

``` R
column_mapping <- c(
  PaO2 = "your_PaO2_variable",
  SpO2 = "your_SpO2_variable",
  FiO2 = "your_FiO2_variable",
  oxygen_supply = "your_oxygen_supply_variable",
  oxygen_mode = "your_oxygen_mode_variable",
  platelets = "your_platelets_variable",
  bilirubin = "your_bilirubin_variable",
  MAP = "your_MAP_variable",
  dopamine = "your_dopamine_variable",
  dobutamine = "your_dobutamine_variable",
  epinephrine = "your_epinephrine_variable",
  norepinephrine = "your_norepineprhine_variable",
  norepinephrine_amp = "your_norepinephrine_amp_variable",
  GCS = "your_GCS_variable",
  creatinine = "your_creatinine_variable"
)
```

Then pass that mapping to the sofa_total function along with your dataframe:

``` R
result <- sofa_total(df, column_mapping)
```
</details>

# Calculations
The SOFA (Sequential Organ Failure Assessment) [@Lambden2019; @Moreno2023] calculation functions are designed to provide the worst SOFA score for a patient within a specific interval.
When retrospectively calculating the SOFA score, one must address missing values and inaccurately recorded values. This function automatically handles such issues.
Consequently, the SOFA score functions incorporate several assumptions to ensure feasible calculations. Users should be aware of these assumptions when utilizing
these functions.

## Respiratory system

+----------------------+---------+---------+---------+-------------+------------+
| Points               | 0       | 1       | 2       | 3           | 4          |
+======================+=========+=========+=========+=============+============+
| PaO~2~/FiO~2~ (mmHg) | ≥400    | <400    | <300    | <200        | <100       |
+----------------------+---------+---------+---------+-------------+------------+
|                      | NOT mechanically ventilated | with respiratory support |
+----------------------+-----------------------------+--------------------------+

Many respiratory parameters tend to be missing. The `sofa_respiration()` function calculated the respiratory points of the SOFA score
with the following rules and assumptions:

### PaO~2~
1. Venous PaO~2~:  
If PaO~2~ is below 7 kPa, it is assumed to be venous and removed from the dataset. The next available value will be used if possible; otherwise, it will be marked as missing. 
2. Low SpO~2~:  
If SpO~2~ is below 50%, the value is removed from the dataset and it is treated as missing [^1].
Note: Pulse oximeters are only calibrated for SpO~2~ values between 70-100%. Values below 70% should be considered qualitatively rather than quantitatively accurate and are therefore set to 70%. [@CHAN2013789]
3. Approximation using SpO~2~:  
When PaO~2~ is unknown (e.g., arterial blood gas not available), it is imputed from SpO~2~ using the Brown's equation [@BROWN2016307]. Rounded to one decimal place. 
Imputation is unreliable for SpO~2~ >96% due to the hemoglobin-oxygen dissociation curve plateau. Therefore, a PaO~2~ of 110 mmHg is used as a substitute for SpO~2~ ≥97% to avoid overestimation. [@severinghaus1979]
4. Missing Values:  
If both PaO~2~ and SpO~2~ are unavailable, the value is treated as missing. Consequently, no points will be assigned to this domain.

### FiO~2~
1. Values Above 100: 
If FiO~2~ ≥101, it is capped at 100 as values above this are not physiologically possible. This issue originates from certain devices where the calculation of FiO~2~ is automated by an algorithm using specific parameters. In some situations, these algorithms may erroneously report values greater than 100, which is technically impossible. To address this, such values are adjusted to a maximum of 100. 
2. FiO~2~ as zero value:  
If FiO~2~ = 0, it is treated as missing.[^1]
3. High FiO~2~ with no other respiratory data:  
If FiO~2~ ≥85 and no oxygen delivery device is known and (Oxygen flow is zero or missing), this is likely an SpO~2~ confusion on data entry. The value is treated as missing.[^1]
4. Fractional FiO~2~ Values:  
If FiO~2~ is below 1, it is assumed to be a fraction and multiplied by 100 to convert it to a percentage. 
5. Low FiO~2~ Values:  
If FiO~2~ is between 1 and 21, it is treated as missing.[^1]
6. Estimation from type of O~2~ delivery and flow rate:
    + If FiO~2~ is missing but oxygen supplementation type and flow rate are available, the FiO~2~ is estimated using the table below. 
    + If only oxygen mode is available, the maximum FiO~2~ for that mode is used. 
    + Estimating FiO~2~ from oxygen flow/delivery rates:

+---------------------------------+-------------------+------------------------------+-------------------------------------------+
| Type of O~2~ delivery           | Flowrate\ (L/min) | FiO~2~                       | Formula                                   |
+=================================+===================+==============================+===========================================+
| Nasal cannula                   | 1 \               | 24%  \                       | $FiO_{2} = 20+(4*Q)$                      |
|                                 | 2 \               | 28%  \                       |                                           |
|                                 | 3 \               | 32%  \                       |                                           |
|                                 | 4 \               | 36%  \                       |                                           |
|                                 | 5 \               | 40%  \                       |                                           |
|                                 | 6 \               | 44%  \                       |                                           |
+---------------------------------+-------------------+------------------------------+-------------------------------------------+
| Non-rebreather mask             | 10-15             | ~60-90%                      | $FiO_{2} = 60 + (Q – 10) * 6$             |
+---------------------------------+-------------------+------------------------------+-------------------------------------------+
| Venturi Mask                    | 2 \               | 24%  \                       | n/a                                       |
|                                 | 4 \               | 28%  \                       |                                           |
|                                 | 6 \               | 31%  \                       |                                           |
|                                 | 8 \               | 35%  \                       |                                           |
|                                 | 10 \              | 40%  \                       |                                           |
|                                 | 15 \              | 60%  \                       |                                           |
+---------------------------------+-------------------+------------------------------+-------------------------------------------+
| High-flow nasal cannula         | Up to 60          | 30-100%                      | $FiO_{2} = 30 + \frac{(Q - 1) * 70}{59}$  |
| (optiflow or cpap)              |                   |                              |                                           |
+---------------------------------+-------------------+------------------------------+-------------------------------------------+


## Coagulation
Calculates the coagulation SOFA-score part based on platelet count (*10^3^ µL^-1^) using the `sofa_coagulation()` function. This section does not correct for unrealistic platelet counts, as this data is generally already curated by the lab.

+--------------------------------+------+-------+-------+------+------+
| Points                         |  0   |  1    |  2    |  3   |  4   |
+================================+======+=======+=======+======+======+
| Platelet Count (*10^3^ µL^-1^) | ≥ 150| < 150 | < 100 | < 50 | < 20 |
+--------------------------------+------+-------+-------+------+------+

## Cardiovascular system
Calculates the SOFA score component for the cardiovascular system using the `sofa_cardiovascular()` function. The first two levels (0 and 1 points) are based on Mean Arterial Pressure (MAP),
while higher levels are determined by catecholamine administration. For dopamine, dobutamine, epinephrine, and norepinephrine, these
dosages need to be provied as µg/kg/min.

+----------+----------------------------+----------------------------------------------------+-----------------------------------------------------------+------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------+
| Points   | 0                          | 1                                                  | 2                                                         | 3                                                                                        | 4                                                                                         |
+==========+============================+====================================================+===========================================================+==========================================================================================+===========================================================================================+
| Criteria | MAP ≥ 70 mmHg              | MAP < 70 mmHg                                      | Dopamine ≤ 5 µg/kg/min OR Dobutamine (any dose)           | Dopamine > 5 µg/kg/min OR Epinephrine ≤ 0.1 µg/kg/min OR Norepinephrine ≤ 0.1 µg/kg/min  | Dopamine > 15 µg/kg/min OR Epinephrine > 0.1 µg/kg/min OR Norepinephrine > 0.1 µg/kg/min  |
+----------+----------------------------+----------------------------------------------------+-----------------------------------------------------------+------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------+

In some cases, norepinephrine is administered as a bolus rather than as a continuous infusion. This is recorded in the `norepinephrine_amp` variable,
which is set to `1` (true) for bolus administration or `0` (false) otherwise. Since the amount of norepinephrine administered differentiates between 3
and 4 points in the cardiovascular SOFA score, the script cannot automatically assign the correct point value. Therefore, **when norepinephrine is
given, 3.5 points are assigned.** Researchers should decide whether to round this value up or down based on their specific research question.

## Liver
Calculates the SOFA score component for the liver using the `sofa_liver()` function. Points are assigned according to the following table. Bilirubin levels should be
entered in μmol/L, mg/dL only for reference.

+----------------------+------+---------+---------+----------+-------+
| Points               |   0  | 1       | 2       | 3        | 4     |
+======================+======+=========+=========+==========+=======+
| Bilirubin (mg/dL)    | <1.2 | 1.2–1.9 | 2.0–5.9 | 6.0–11.9 | ≥12.0 |
+----------------------+------+---------+---------+----------+-------+
| Bilirubin (μmol/L)   |  <20 | 20–32   | 33–101  | 102–204  |  ≥205 |
+----------------------+------+---------+---------+----------+-------+

## Central Nervous System
Calculates the SOFA score component for the central nervous system using the `sofa_cns()` function. Points are assigned based on the Glascow Coma Scale (GCS).

+----------------------------+----+-------+-------+-----+----+
| Points                     | 0  | 1     | 2     | 3   | 4  |
+============================+====+=======+=======+=====+====+
| Glasgow Coma Scale (GCS)   | 15 | 13–14 | 10–12 | 6–9 | <6 |
+----------------------------+----+-------+-------+-----+----+

A GCS score < 3 reflects severe neurological dysfunction and is assigned the maximum score, eg in patient with a tube.

## Renal
Calculates the SOFA score compontent for the renal function using the `sofa_renal()` function. Points are assigned based on the creatinine level (µmol/L).
This section does not correct for unrealistic creatinine levels, as this data is generally already curated by the lab.

The following table lists the assigned points: [@Vincent1996]
Creatine should be provided in µmol/L, mg/dL for reference.

+------------------------+---------+---------+---------+---------+---------+
| Points                 | 0       | 1       | 2       | 3       | 4       |
+========================+=========+=========+=========+=========+=========+
| Creatinine (mg/dL)     | < 1.2   | 1.2–1.9 | 2.0–3.4 | 3.5–4.9 | ≥ 5.0   |
+------------------------+---------+---------+---------+---------+---------+
| Creatinine (μmol/L)    | < 110   | 110–170 | 171–299 | 300–440 | ≥ 440   |
+------------------------+---------+---------+---------+---------+---------+
| Urine Output (mL/day)  | \> 500  |         |         | < 500   | < 200   |
+------------------------+---------+---------+---------+---------+---------+

The following assumptions are made when calculating the renal SOFA score:

Urine output < 500 mL/day correlates with moderate renal dysfunction, while < 200 mL/day indicates severe dysfunction. Urine output is currently not implemented in the calculation, as this data is often missing or registred inadequately.

[^1]: When extracted from the Acutelines database the next available value is used if possible, otherwise it is marked as missing.