---
title: "SOFA score calculation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SOFA-score}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, eval = FALSE}
library(acutelines.datatoolbox)
```

## Usage
Asuming the data is loaded into a dataframe `df`, the SOFA score can be calculated in two different ways.

### Automatic calculation
Use the sofa_magic_wrapper() function when you use a raw data file from the Acutelines database. SOFA related variables are coded as `SOFA_[timespan]h_[variable]`. Make sure all variables are present.
`interval` can be used to set the time stap interval in correspondance to your dataset, while `timespan` can be use to set the maximum timespan.

    result <- sofa_magic_wrapper(df, interval=24, timespan=72)

### Manual calculation
When your dataset has a different naming schema (not having the `SOFA_[timespan]h_[variable]` format), you can still calculate the SOFA score with this package, however, you have to
provide the package with a mapping of each column in your dataset corresponding to the acutal element of the SOFA score.

First define the mapping:

    column_mapping <- c(
      PaO2 = "your_PaO2_variable",
      SpO2 = "your_SpO2_variable",
      FiO2 = "your_FiO2_variable",
      oxygen_supply = "your_oxygen_supply_variable",
      platelets = "your_platelets_variable",
      bilirubin = "your_bilirubin_variable",
      MAP = "your_MAP_variable",
      dopamine = "your_dopamine_variable",
      dobutamine = "your_dobutamine_variable",
      epinephrine = "your_epinephrine_variable",
      norepinephrine = "your_norepineprhine_variable",
      norepinephrine_amp = "your_norepinephrine_amp_variable",
      GCS = "your_GCS_variable",
      creatinine = "your_creatinine_variable"
    )

Than pass that mapping to the sofa_total function along with your dataframe:

    result <- sofa_total(df, column_mapping)

## Calculations
The SOFA (Sequential Organ Failure Assessment) calculation functions are designed to provide the worst SOFA score for a patient within a specific interval.
When retrospectively calculating the SOFA score, one must address missing values and inaccurately recorded values. This function automatically handles such issues.
Consequently, the SOFA score functions incorporate several assumptions to ensure feasible calculations. Users should be aware of these assumptions when utilizing
these functions.

### Respiration

### Coagulation

### Cardiovascular

### Central nervous system

### Renal

## References
1. Lambden, S., Laterre, P.F., Levy, M.M. *et al.* The SOFA score—development, utility and challenges of accurate assessment in clinical trials. *Crit Care* **23**, 374 (2019). https://doi.org/10.1186/s13054-019-2663-7
2. Moreno, R., Rhodes, A., Piquilloud, L. *et al.* The Sequential Organ Failure Assessment (SOFA) Score: has the time come for an update?. *Crit Care* **27**, 15 (2023). https://doi.org/10.1186/s13054-022-04290-9